---
execute:
  echo: true
  eval: true
format:
  html:
    suppress-bibliography: true
---

# R Session Info {.unnumbered}

This document lists all R packages used across analysis notebooks in this repository, and details how to reproduce or create similar analysis environments.

## Package Loading

```{r}
#| warning: FALSE
#| message: FALSE

library(tidyverse)      # Data manipulation and ggplot2
library(ggtext)         # Markdown text in ggplot2
library(viridis)        # Color palettes
library(RColorBrewer)   # Color palettes
library(knitr)          # Dynamic report generation
library(ggh4x)          # Extensions for ggplot2
library(cowplot)        # Plot arrangement
library(png)            # Reading PNG images
library(grid)           # Grid graphics
library(reshape2)       # Data reshaping
library(seqinr)         # Reading and writing FASTA files
```

## Session Information

This shows the exact versions of R and all packages used in the analysis.

```{r}
sessionInfo()
```

## Reproducibility and Environment

This repository manages R package versions using [`renv`](https://rstudio.github.io/renv/).

To reproduce the environment for this project:

```{r}
#renv::init() # Was used to create the initial renv environment and lockfile
#renv::snapshot() # Was used to update the renv.lock file after installing new packages
#renv::restore() # Restores the R environment based on the renv.lock file
```

This command uses the `renv.lock` file to install the exact package versions required for the analysis.