---
execute:
  eval: FALSE
  message: FALSE
  warning: FALSE
---

# Anvio {.unnumbered}

```{r}
library(tidyverse)
library(ggtext)
library(viridis)
library(RColorBrewer)
```

## Set up paths

```{r}
#This will open the .Renviron file for the project. You need to add a DATA_PATH environment variable with the path to your data folder.
#usethis::edit_r_environ(scope = "project") 

# This will create and open the .bash_env file for editing. You need to add a DATA_PATH environment variable with the path to your data folder.
#file.create(".bash_env")
#file.edit(".bash_env") 

FOLDER_PATH <- Sys.getenv("DATA_PATH") 
genomes_list_folder <- file.path(FOLDER_PATH, "genome_lists")
anvio_folder <- file.path(FOLDER_PATH, "anvio-9")
```

This notebook contains analyses using anvi'o version 9 [@eren2020; @delmont2018]. For more in-depth information about anvi'o go to their official online site here: <https://anvio.org/>

We first source the `.bash_env` file to set up the environment variables.

```{bash, eval=FALSE}
source .bash_env
```

## Functional annotations

We use as input (`path_i`) the anvi'o contigs database files (`.db`) that were generated in [[Methods_Prokka_Annotations]{.underline}](Methods_Prokka_Annotations.html).

### COG annotation

First we downloaded and set up the NCBI's Clusters of Orthologous Groups database[@Tatusov2000, @Galperin2021, @galperin2024] using `anvi-setup-ncbi-cogs`. The default is the latest version, which is COG24, meaning that anvi'o will use the NCBI's 2024 release of COGs to setup the database.

The `anvi-run-ncbi-cogs` command was used to annotate the contig databases. We used DIAMOND as searching method[@buchfink2014] with the recommended "sensitive" option.

```{bash, eval=FALSE}
conda deactivate
conda activate anvio-9

path_i="$DATA_PATH/anvio-9/contigs_db"

for file in $path_i/*.db; do
    anvi-run-ncbi-cogs -T 8 --search-with diamond -c $file;
done
```

### KEGG annotation

First we downloaded and set up the KEGG KOfam[@Kanehisa2000; @Kanehisa2023] database using `anvi-setup-kegg-data`. We used the snapshot of the KEGG database dated 2025-11-07 (hash 68221bd12b3).

The program `anvi-run-kegg-kofams` annotates genes in a given anvi'o contigs database with KEGG Orthology (KO) numbers via hits to the KEGG KOfam database [@aramaki2019]. We used `-E` lower than 1e-15. 

```{bash, eval=FALSE}
conda deactivate
conda activate anvio-9

path_i="$DATA_PATH/anvio-9/contigs_db"

for file in $path_i/*.db; do
    anvi-run-kegg-kofams -T 8 --log-bitscores -E 1e-15 --just-do-it -c $file;
done
```

The option `--log-bitscores` records the bit scores values. They get saved to the main folder for the repo, so we moved them to a new subfolder.

```{r}
path_i <- "."
path_o <- file.path(anvio_folder, "KEGG_bitscores")
dir.create(path_o, recursive = TRUE, showWarnings = FALSE)
files_to_move <- list.files(path_i, pattern = "_bitscores.txt$", full.names = TRUE)

for (file in files_to_move) {
  new_path <- file.path(path_o, basename(file))
  print(new_path)
  file.rename(file, new_path)
}
```

### PFAM annotation

First you need to download and set up the PFAM database[@Bateman2000; @Mistry2021] using `anvi-setup-pfams`: this is something you will do only once. We used Pfam version 38.1 (2025-09)

The program `anvi-run-kegg-pfam` annotates genes in a given anvi'o contigs database with the PFAM database.

```{bash, eval=FALSE}
conda deactivate
conda activate anvio-9

path_i="$DATA_PATH/anvio-9/contigs_db"

for file in $path_i/*.db; do
    anvi-run-pfams -T 8 -c $file;
done
```

### CAZyme Annotation

First you need to download and set up the CAZyme database [@drula2021] using `anvi-setup-cazymes`: this is something you will do only once. We used Pfam version 12.

The program `anvi-run-cazymes` annotates genes in your contigs-db with the dbCAN CAZyme database.

```{bash, eval=FALSE}
conda deactivate
conda activate anvio-9

path_i="$DATA_PATH/anvio-9/contigs_db"

for file in $path_i/*.db; do
    anvi-run-cazymes -T 8 -c $file;
done
```

### Getting HMMs

When you run the `anvi-run-hmms` command the occurrence of bacterial single-copy genes across your contigs is added to your contigs database [@Eddy2011].

```{bash, eval=FALSE}
conda deactivate
conda activate anvio-9

path_i="$DATA_PATH/anvio-9/contigs_db"

for file in $path_i/*.db; do
    anvi-run-hmms -T 8 -c $file --just-do-it;
done
```

### Getting Metabolic Networks

The program `anvi-reaction-network` is used to store a metabolic reaction-network in each contigs-db. The network consists of data on biochemical reactions predicted to be encoded by each genome, referencing the KEGG Orthology (KO) and ModelSEED Biochemistry databases.

```{bash, eval=FALSE}
conda deactivate
conda activate anvio-9

path_i="$DATA_PATH/anvio-9/contigs_db"

for file in $path_i/*.db; do
    anvi-reaction-network -c $file;
done
```

### Annotation Summaries

This generates basic stats on the annotations:

```{bash, eval=FALSE}
conda deactivate
conda activate anvio-9

path_i="$DATA_PATH/anvio-9/contigs_db"
path_o="$DATA_PATH/anvio-9/annotation_summaries"
mkdir -p "$path_o"

for file in $path_i/*.db; do
    FILENAME=`basename ${file%.*}`
    anvi-display-contigs-stats $file --report-as-text -o $path_o/stats_$FILENAME.txt;
done
```

This provides basic info on the contigs databases:

```{bash, eval=FALSE}
conda deactivate
conda activate anvio-9

path_i="$DATA_PATH/anvio-9/contigs_db"
path_o="$DATA_PATH/anvio-9/annotation_summaries"

for file in $path_i/*.db; do
    anvi-db-info $file &>> $path_o/anvi-db-info_log.txt;
done
```

## Metabolic Analysis

### Estimation of KEGG module completeness

Using `anvi-estimate-metabolism` we predicted KEGG module completeness as described in [@veseli2023] for the the organisms listed in `NovCor_MetabolicAnalysis_GenomeList_v01.txt`. This list includes the 30 *Corynebacterium* genomes selected for metabolic analysis.

```{bash, eval=FALSE}
#conda activate anvio-dev

txt_file="data/genome_lists/NovCor_MetabolicAnalysis_GenomeList_v03.txt"
prefix=NovCor
path_o="data/Anvio8/Metabolic_Analysis"
mkdir -p "$path_o"

anvi-estimate-metabolism -e $txt_file \
                         -O $path_o/$prefix --add-copy-number --output-modes hits,modules
```

The generated `NovCor_modules.txt` and `NovCor_hits.txt` files are included as Supp Tables in this publication.

#### Code for KEGG Manuscript Table

```{r}
# Read modules data and make appropriate labeling columns
Modules <- read_delim("data/Anvio8/Metabolic_Analysis/NovCor_modules.txt", show_col_types = FALSE, col_types = cols(unique_enzymes_hit_counts = col_character(), gene_caller_ids_in_module = col_character())) %>% 
  mutate(module_class_short = paste0("(", substr(module_class, 1, 1), ")")) %>%
  unite(modules_label, c(module_class_short, module_category, module_subcategory), sep = "_", remove = FALSE) %>% 
  mutate_at(c("genome_name", "module", "module_name", "module_category", "module_subcategory", "modules_label"), factor)

# Summary tally table of fully stepwise complete modules by species. Filter to modules present in at least 25 genomes
ModulesTable <- Modules %>%
  filter(stepwise_module_completeness == 1) %>%
  group_by(module_class, module_category, module_subcategory, module, module_name) %>%
  tally() %>%
  ungroup() %>%
  select(module, module_name, n, -module_class) %>%  
  arrange(desc(n)) %>%
  filter(n > 25)

write_csv(ModulesTable, "data/Anvio8/Metabolic_Analysis/Fig_Tables/NovCor_KEEG_Table.csv")
kable(ModulesTable)
```

#### Code for KEGG Manuscript Figure

```{r}
StrainOrder <- read_csv("data/genome_lists/NovCor_MetabolicAnalysis_PlotOrder.txt")

# When using stepwise_module_completeness some modules can be = 0 for all genomes but still have pathwise_module_completeness > 0. We are removing those modules.
Modules_cleaned <- Modules %>%
  group_by(module) %>%
  filter(!all(stepwise_module_completeness == 0)) %>%
  ungroup() %>%
  mutate(across(where(is.factor), droplevels))

all_zero_modules <- setdiff(Modules$module, Modules_cleaned$module)
all_zero_subcategory <- setdiff(Modules$module_subcategory, Modules_cleaned$module_subcategory)

# Calculating mean_stepwise_module_completeness and preparing plot labels 
Modules_Means = Modules_cleaned %>% 
  group_by(modules_label, genome_name, .drop = FALSE) %>%
  summarise(mean_stepwise_module_completeness = mean(stepwise_module_completeness)) %>%
  mutate(mean_stepwise_module_completeness = coalesce(mean_stepwise_module_completeness, 0)) %>%
  mutate(species_name = str_extract(genome_name, "^[^_]+")) %>%
  mutate(module_category = str_extract(modules_label, "^[^_]+_[^_]+")) %>%
  mutate(module_category = str_replace_all(module_category, "_", " ")) %>%
  mutate(modules_final_label = str_replace(modules_label, "^[^_]+_[^_]+_", "")) %>%
  mutate_at(c("module_category", "modules_final_label"), factor) 

Modules_Means <- left_join(Modules_Means, StrainOrder, by = "genome_name")

Modules_Means$species_name <- sub("Cke", "Cna", Modules_Means$species_name)
Modules_Means$species_name <- sub("Cau", "Cmq", Modules_Means$species_name)
Modules_Means$species_name <- factor(Modules_Means$species_name, levels = c("Cna","Ccu","Cha","Cyo","Cmq","Ctu"))
Modules_Means$genome_name <- factor(Modules_Means$genome_name, levels = StrainOrder$genome_name)
Modules_Means$styled_genome_name <- factor(Modules_Means$styled_genome_name, levels = rev(StrainOrder$styled_genome_name))
Modules_Means$modules_final_label  <- factor(Modules_Means$modules_final_label, rev(levels(Modules_Means$modules_final_label)))
```

```{r}
#| fig-width: 12
#| fig-height: 13

# Colors for plot
colorsSpecies <- c("#8200E7", "#FF00FF", "#0ABF00", "#0A6003", "#39D9E5", "#002BFF")

# Heatmap plot
plot <- ggplot(Modules_Means, aes(x = styled_genome_name, y = modules_final_label, fill = mean_stepwise_module_completeness)) + 
  geom_tile(color = "white") +
  facet_grid2(module_category ~ species_name, scales = "free", space = "free", switch = "y", labeller = label_wrap_gen(width = 31),
              #strip = strip_themed(background_x = elem_list_rect(fill = colorsSpecies))) +
              strip = strip_themed(text_x = elem_list_text(colour = colorsSpecies))) +
  labs(x = "", y = "") +
  scale_y_discrete(expand = c(0,0), position = "right") +
  scale_x_discrete(expand = c(0,0)) +
  scale_fill_viridis(direction = -1, option = "A", begin = 0.2) +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        strip.background = element_rect(colour = "black", fill = NA),
        strip.text.y.left = element_text(angle = 0, hjust = 0.5, size = 11),
        strip.text.x = element_text(size = 12, face = "bold"),
        axis.text.y = element_text(size = 11, color = "black"),
        axis.text.x = element_markdown(angle = 45, hjust = 1, size = 11, color = "black"),
        panel.spacing = unit(0, 'pt'),
        legend.position = c(1.30, -0.11), 
        legend.direction = "horizontal",
        legend.justification = c(1, 0)) +
  guides(fill = guide_colourbar(title = "Average stepwise\nmodule completeness", title.position = "top", title.hjust = 0.5, barwidth = 8))


ggsave(plot = plot, filename = "data/Anvio8/Metabolic_Analysis/Fig_Tables/NovCor_KEEG_Figure.png", height = 13, width = 12, device = 'png', dpi = 600)

plot
```

### Module Enrichment Analysis

The `NovCor_modules.txt` file generated in the previous step is used by `anvi-compute-metabolic-enrichment` to calculate KEGG module enrichment[@shaiber2020] across different groups of genomes. We used the `-G` flag to indicate the grouping options defined in file `NovCor_MetabolicAnalysis_Groups_CtuComplex.txt`that includes the 30 *Corynebacterium* genomes grouped by their final species assignment.

```{bash, eval=FALSE}
#conda activate anvio-dev

csv_file_modules="data/Anvio8/Metabolic_Analysis/NovCor_modules.txt"
path_o="data/Anvio8/Metabolic_Analysis/Enrichment_Modules"
mkdir -p "$path_o"

txt_file_groups="data/genome_lists/NovCor_MetabolicAnalysis_Groups_CtuComplex_v03.txt"
output_file="$path_o/NovCor_enriched_modules_stepwise_1_CtuComplex.txt"
anvi-compute-metabolic-enrichment -M "$csv_file_modules" \
                                  -G "$txt_file_groups" \
                                  -o "$output_file" \
                                  --use-stepwise-completeness \
                                  --module-completion-threshold 1.0
                                  
txt_file_groups="data/genome_lists/NovCor_MetabolicAnalysis_Groups_Cna.txt"
output_file="$path_o/NovCor_enriched_modules_stepwise_1_Cna.txt"
anvi-compute-metabolic-enrichment -M "$csv_file_modules" \
                                  -G "$txt_file_groups" \
                                  -o "$output_file" \
                                  --use-stepwise-completeness \
                                  --module-completion-threshold 1.0      

txt_file_groups="data/genome_lists/NovCor_MetabolicAnalysis_Groups_Cha.txt"
output_file="$path_o/NovCor_enriched_modules_stepwise_1_Cha.txt"
anvi-compute-metabolic-enrichment -M "$csv_file_modules" \
                                  -G "$txt_file_groups" \
                                  -o "$output_file" \
                                  --use-stepwise-completeness \
                                  --module-completion-threshold 1.0 

```

### KO Enrichment Analysis

We use `anvi-display-functions`[@shaiber2020]. It runs the enrichment for the desired annotations (in this case KOfam). It saves the output to a temp file that can be recovered, renamed and saved into the same `path_o` as the database file.

```{bash, eval=FALSE}
#conda activate anvio-dev

txt_file="data/genome_lists/NovCor_MetabolicAnalysis_GenomeList_v01.txt"
path_o="data/Anvio8/Metabolic_Analysis/Enrichment_KOs"
mkdir -p "$path_o"

txt_file_groups="data/genome_lists/NovCor_MetabolicAnalysis_Groups_CtuComplex_v03.txt"
anvi-display-functions -e $txt_file \
                       --groups-txt $txt_file_groups \
                       --annotation-source KOfam \
                       --profile-db $path_o/KOfam_CtuComplex.db

txt_file_groups="data/genome_lists/NovCor_MetabolicAnalysis_Groups_Cna.txt"
anvi-display-functions -e $txt_file \
                       --groups-txt $txt_file_groups \
                       --annotation-source KOfam \
                       --profile-db $path_o/KOfam_Cna.db
                       
txt_file_groups="data/genome_lists/NovCor_MetabolicAnalysis_Groups_Cha.txt"
anvi-display-functions -e $txt_file \
                       --groups-txt $txt_file_groups \
                       --annotation-source KOfam \
                       --profile-db $path_o/KOfam_Cha.db
```

Many of the enriched KOs belong to the already identified enriched modules. This function filters out those KOs that are part of the definition of modules listed in the enriched modules identified with `anvi-compute-metabolic-enrichment`:

```{r, eval=FALSE}
get_KOs_not_in_enrich_modules <- function(modules, hits, enrich_modules, enrich_KOs, q_value_threshold) {

  # Get list of KOs that correspond with the modules on the enrichment file
  modules_subset <- modules %>% filter(module %in% enrich_modules$accession)
  module_definition_subset <- paste(modules_subset$module_definition, collapse = " ")
  module_definition_subset <- str_extract_all(module_definition_subset, "\\bK\\d+\\b")
  module_definition_subset <- unique(unlist(module_definition_subset))

  # Add a column to enrich_KOs indicating if the KO is the previous list
  enrich_KOs$inModule <- sapply(strsplit(as.character(enrich_KOs$accession), ","), function(x) all((x %in% module_definition_subset)))

  # Filter Enrich_KO based on inModule and q_value_threshold
  enrich_KOs_filtered <- enrich_KOs %>% 
    filter(inModule == FALSE) %>% 
    filter(adjusted_q_value < q_value_threshold)

  return(enrich_KOs_filtered)
}

Modules <- read_delim("data/Anvio8/Metabolic_Analysis/NovCor_modules.txt", show_col_types = FALSE, col_types = cols(unique_enzymes_hit_counts = col_character(), gene_caller_ids_in_module = col_character()))
KOHits <- read_delim("data/Anvio8/Metabolic_Analysis/NovCor_hits.txt", show_col_types = FALSE)

Enrich_Mod_CtuComplex <- read_delim("data/Anvio8/Metabolic_Analysis/Enrichment_Modules/NovCor_enriched_modules_stepwise_1_CtuComplex.txt", show_col_types = FALSE)
Enrich_KO_CtuComplex <- read_delim("data/Anvio8/Metabolic_Analysis/Enrichment_KOs/NovCor_enriched_KOs_CtuComplex.txt", show_col_types = FALSE)
Enrich_KO_CtuComplex_Filtered <- get_KOs_not_in_enrich_modules(Modules, KOHits, Enrich_Mod_CtuComplex, Enrich_KO_CtuComplex, 0.1)
write_csv(Enrich_KO_CtuComplex_Filtered, "data/Anvio8/Metabolic_Analysis/Enrichment_KOs/NovCor_enriched_KOs_CtuComplex_Filtered.csv")
```

```{r, eval=FALSE}
Enrich_Mod_CtuComplex <- read_delim("data/Anvio8/Metabolic_Analysis/Enrichment_Modules/NovCor_enriched_modules_stepwise_1_CtuComplex.txt", show_col_types = FALSE)
Enrich_KO_CtuComplex <- read_delim("data/Anvio8/Metabolic_Analysis/Enrichment_KOs/NovCor_enriched_KOs_CtuComplex.txt", show_col_types = FALSE)

Enrich_Mod_Cna <- read_delim("data/Anvio8/Metabolic_Analysis/Enrichment_Modules/NovCor_enriched_modules_stepwise_1_Cna.txt", show_col_types = FALSE)
Enrich_KO_Cna <- read_delim("data/Anvio8/Metabolic_Analysis/Enrichment_KOs/NovCor_enriched_KOs_Cna.txt", show_col_types = FALSE)

Enrich_Mod_Cha <- read_delim("data/Anvio8/Metabolic_Analysis/Enrichment_Modules/NovCor_enriched_modules_stepwise_1_Cna.txt", show_col_types = FALSE)
Enrich_KO_Cha <- read_delim("data/Anvio8/Metabolic_Analysis/Enrichment_KOs/NovCor_enriched_KOs_Cna.txt", show_col_types = FALSE)
```

### Pangenomic level analysis

This analysis using `NovCor_MetabolicAnalysis_GenomeList_v02.txt` includes the 30 *Corynebacterium* genomes selected for metabolic analysis.

```{bash, eval=FALSE}
#conda activate anvio-dev

path_i="data/Anvio8/Pangenomic_Analysis"
mkdir -p "$path_i"

anvi-gen-genomes-storage -e "data/genome_lists/NovCor_MetabolicAnalysis_GenomeList_v02.txt" -o $path_i/NovCor-ANI-GENOMES.db --gene-caller Prodigal

anvi-db-info $path_i/NovCor-ANI-GENOMES.db

anvi-pan-genome -g $path_i/NovCor-ANI-GENOMES.db \
                -o $path_i/ANI -n PAN_ANI \
                --use-ncbi-blast --mcl-inflation 10  --num-threads 8

#Create a bin collection named ANI with all Core GCs for the species using the visual interface:       

anvi-display-pan -p $path_i/ANI/PAN_ANI-PAN.db -g $path_i/NovCor-ANI-GENOMES.db         

anvi-summarize -p $path_i/ANI/PAN_ANI-PAN.db \
               -g $path_i/NovCor-ANI-GENOMES.db \
               -o $path_i/ANI/Summaries \
               -C ANI
```
